---
title: "Applying a bayesian workflow: lessons from Statistical Rethinking (I)"
output: html_notebook
---

### 1) Exploratory data analysis and data transformation


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rstan)
library(bayesplot)

# Detect cores for parallel sampling
options(mc.cores = parallel::detectCores())

df = read_csv('../data/raw/train.csv') %>% select('SalePrice', 'LotArea', 'Neighborhood')

head(df)
```

```{r}
# Perform log transfomations
df <- df %>% mutate(log_sales_price = log(SalePrice),
                    log_lot_area = log(LotArea),
                    neighbourhood = as.integer(as.factor(df$Neighborhood)))
```

```{r}
ggplot(df, aes(x = log_lot_area, y = log_sales_price)) +
  geom_point(colour = 'blue') +
  geom_smooth(method = lm, se = FALSE, formula = 'y ~ x') + 
  ggsave('figures/2r_pooling_scatter.svg', dpi = 72, width=10, height = 8, units = 'in')
```


```{r}
ggplot(df, aes(x = log_lot_area, y = log_sales_price)) +
  geom_point(colour = 'blue') +
  geom_smooth(method = lm, se = FALSE, formula = 'y ~ x', fullrange = TRUE) +
  facet_wrap(~Neighborhood) +
  theme(strip.background = element_blank()) + 
  ggsave('figures/3r_facet_scatter.svg', dpi = 72, height = 8, width = 7)
```


```{r}
df = df %>% mutate(log_lot_area_z = scale(log_lot_area),
                    log_sales_price_z = scale(log_sales_price))
```

```{r}
df %>% select(log_sales_price_z) %>% 
  apply(., MARGIN  = 2, FUN = function(x) exp((x * sd(df$log_sales_price)) + mean(df$log_sales_price))) %>% 
  as.data.frame() %>% 
  cbind(., df$SalePrice) %>% 
  mutate(delta = log_sales_price_z - `df$SalePrice`) %>% 
  round(., digits = 1) %>% 
  summarise(delta = sum(delta))

```


$$
y_i \sim Normal(\mu_i, \sigma) \\
\mu_i = \alpha_{j} + \beta * x_i \\
\alpha_j\sim Normal(0, 1)\\
\beta\sim Normal(0, 1) \\
\sigma\sim exp(1)
$$




### 3)  Prior predictive checks - simulate data from the implied generative model

```{r include=FALSE}
no_pooling_stan_code = "
// No pooling model for predicting housing prices
data {
    // Fitting the model on training data
    int<lower=0> N; // Number of rows
    int<lower=0> neighbourhood[N]; // neighbourhood categorical variable
    int<lower=0> N_neighbourhood; // number of neighbourhood categories
    vector[N] log_sales_price; // log sales price
    vector[N] log_lot_area; // log lot area

    // Adjust scale parameters in python
    real alpha_sd;
    real beta_sd;
    
    // Set to zero for prior predictive checks, set to one to evaluate likelihood
    int<lower = 0, upper = 1> run_estimation;
}
parameters {
    vector[N_neighbourhood] alpha; // Vector of alpha coefficients for each neighbourhood
    real beta;
    real<lower=0> sigma;
}
model {
    // Priors
    target += normal_lpdf(alpha | 0, alpha_sd);
    target += normal_lpdf(beta | 0, beta_sd);
    target += exponential_lpdf(sigma |1);
    //target += normal_lpdf(sigma |0, 1);
    
    // Likelihood
    if(run_estimation==1){
        target += normal_lpdf(log_sales_price | alpha[neighbourhood] + beta * log_lot_area, sigma);

    }
}
generated quantities {
    // Uses fitted model to generate values of interest without re running the sampler
    vector[N] log_lik; // Log likelihood
    vector[N] y_hat; // Predictions using training data
    {
    for(n in 1:N){
          log_lik[n] = normal_lpdf(log_sales_price | alpha[neighbourhood[n]] + beta * log_lot_area[n], sigma);
          y_hat[n] = normal_rng(alpha[neighbourhood[n]] + beta * log_lot_area[n], sigma);      
        }
    }
}
"

# List contains all data inputs
npm_data_check = list(N = nrow(df),
                      log_sales_price = as.vector(df$log_sales_price_z),
                      log_lot_area = as.vector(df$log_lot_area_z),
                      neighbourhood = as.vector(df$neighbourhood),
                      N_neighbourhood = max(df$neighbourhood),
                      alpha_sd = 1, 
                      beta_sd = 1, 
                      run_estimation = 0)

# Draw samples from joint prior distribution
fit_npm_check = stan(model_code = no_pooling_stan_code, data = npm_data_check, chains = 4, seed = 12345)

# Extract samples into a pandas dataframe
npm_df_check = as.data.frame(fit_npm_check)
```

```{r include=FALSE}
# Fit model with diffuse priors
npm_data_check_wide = list(N = nrow(df),
                      log_sales_price = as.vector(df$log_sales_price_z),
                      log_lot_area = as.vector(df$log_lot_area_z),
                      neighbourhood = as.vector(df$neighbourhood),
                      N_neighbourhood = max(df$neighbourhood),
                      alpha_sd = 10, 
                      beta_sd = 10, 
                      run_estimation = 0)

fit_npm_check_wide = stan(model_code = no_pooling_stan_code, data=npm_data_check_wide, chains = 4, seed = 12345)
npm_df_check_wide = as.data.frame(fit_npm_check_wide)
```

```{r}
# Create length of std x variables
x <- seq(from = -3, to = 3, length.out = 200)

# Create empty dataframe and fill it with parameters
df_wide <- as.data.frame(matrix(ncol=100, nrow=200))
for (i in 1:100) {
  alpha <- npm_df_check_wide$`alpha[4]`[i]
  beta <- npm_df_check_wide$beta[i]
  df_wide[, i] <- alpha + beta * x
  
}

# Tidy up filled dataframe
df_wide <- df_wide %>% mutate(x = x) %>% pivot_longer(starts_with("V"))

# Plot
ggplot(df_wide, aes(x = x, y = value)) +
  geom_line(aes(group = name), size = 0.2) +
  scale_x_continuous(breaks = seq(-3, 3, 1)) +
  labs(title = 'Prior predictive checks -- Uninformative (flat) priors',
             x = 'x (z-scores)',
             y = 'Fitted y (z_scores)') + 
  ggsave('figures/5r_prior_predictive_check_wide.svg', dpi = 72, height = 4, width = 7)

```



```{r}
# Create length of std x variables
x <- seq(from = -3, to = 3, length.out = 200)

# Create empty dataframe and fill it with parameters
df_regularising <- as.data.frame(matrix(ncol=100, nrow=200))
for (i in 1:100) {
  alpha <- npm_df_check$`alpha[4]`[i]
  beta <- npm_df_check$beta[i]
  df_regularising[, i] <- alpha + beta * x
  
}

# Tidy up filled dataframe
df_regularising <- df_regularising %>% mutate(x = x) %>% pivot_longer(starts_with("V"))

# Plot
ggplot(df_regularising, aes(x = x, y = value)) +
  geom_line(aes(group = name), size = 0.2, colour = 'blue') +
  scale_x_continuous(breaks = seq(-3, 3, 1)) +
  labs(title = 'Prior predictive checks -- Weakly regularizing priors',
             x = 'x (z-scores)',
             y = 'Fitted y (z_scores)') + 
  ggsave('figures/4r_prior_predictive_check.svg', dpi = 72, height = 4, width = 7)
```

```{r}
ggplot(df_wide, aes(x = x, y = value)) +
  geom_line(aes(group = name), size = 0.2) +
  geom_line(data = df_regularising, aes(group = name), size = 0.2, colour = 'blue') +
  scale_x_continuous(breaks = seq(-3, 3, 1)) +
  labs(title = 'Prior predictive checks -- Uninformative (black) vs weakly informative (blue)',
             x = 'x (z-scores)',
             y = 'Fitted y (z_scores)') + 
  ggsave('figures/5r_prior_predictive_check_compare.svg', dpi = 72, height = 4, width = 7)
```



### 4)  Fit model on fake data

```{r include=FALSE}
# Pick random simulation, let's say 50
random_draw <- 50

# Extract the simulated (fake) data implied by the parameters in sample 50
y_sim <-  npm_df_check[random_draw, ] %>% select(contains('y_hat')) %>% t()

# Extract the parameteres corresponding to sample 50
true_parameters = npm_df_check[random_draw,] %>% select(contains(c('alpha','beta','sigma')))

# Fit the model on the fake data
npm_data_check_ = list(N = nrow(df),
                      log_sales_price = as.vector(y_sim), # target is now extracted fake data in sample 50
                      log_lot_area = as.vector(df$log_lot_area_z),
                      neighbourhood = as.vector(df$neighbourhood),
                      N_neighbourhood = max(df$neighbourhood),
                      alpha_sd = 1, 
                      beta_sd = 1, 
                      run_estimation = 1)

fit_npm_check_ = stan(model_code = no_pooling_stan_code, data=npm_data_check_, chains = 4, seed = 12345)
npm_df_check_ = as.data.frame(fit_npm_check_)

fake_fit = npm_df_check_ %>% select(contains(c('alpha', 'beta', 'sigma')))
parameter_df = fake_fit %>% pivot_longer(everything()) %>% rename(parameters = name)
parameter_df$parameters <- factor(parameter_df$parameters, levels = (parameter_df$parameters %>% unique()))

ggplot(parameter_df, aes(value)) + 
  geom_density(colour = 'blue') + 
  facet_wrap(~parameters, scales = 'free') + 
  geom_vline(data = (true_parameters %>% pivot_longer(everything()) %>% rename(parameters = name)), aes(xintercept = value), colour = 'red') + 
  labs(title = 'Model Checking - red lines are "true" parameters',
       x = '') + 
  theme(strip.background = element_blank()) + 
  ggsave('figures/6r_fit_fake_data.svg', dpi = 72, height = 6, width = 9)
```

### 5)  Estimate model on real data

Set `run_estimation=1` and run the code to fit the model. Stan will sample the joint posterior distribution using the default Markov chain Monte Carlo (MCMC) algorithm, the [No-U-Turn sampler (NUTs)](https://mc-stan.org/docs/2_18/reference-manual/hmc-algorithm-parameters.html).

```{r include=FALSE}
# Dictionary with data inputs - set run_estimation=1
npm_data = list(N = nrow(df),
                      log_sales_price = as.vector(df$log_sales_price_z),
                      log_lot_area = as.vector(df$log_lot_area_z),
                      neighbourhood = as.vector(df$neighbourhood),
                      N_neighbourhood = max(df$neighbourhood),
                      alpha_sd = 1, 
                      beta_sd = 1, 
                      run_estimation = 1)

# Fit model by sampling from posterior distribution
fit_npm = stan(model_code = no_pooling_stan_code, data = npm_data, chains = 4, seed = 12345)

# Extract samples into dataframe
fit_npm_df = as.data.frame(fit_npm)
```

### 6) Check whether MCMC sampler and model fit
Stan won't have trouble sampling from such a simple model, so I won't go through chain diagnostics. However, we can see the posterior distributions of all the parameters by looking at the traceplot. I've included number of effective samples and Rhat diagnostics for completeness.

```{r}
stan_plot(fit_npm, 
          show_density = FALSE, 
          unconstrain = TRUE, 
          pars = c('alpha', 'beta', 'sigma')) + 
  labs(title = 'Posterior distributions of fitted parameters') +
  ggsave('figures/7r_posterior.svg', dpi = 72, height = 6, width = 9)
```
<center>
![](figures/7r_posterior.svg)
</center>

```{r echo=FALSE}
# Inspect model fit
color_scheme_set("mix-blue-red")

trace <- mcmc_combo(
 as.array(fit_npm),
 combo = c("dens_overlay", "trace"),
 pars = c('alpha[1]', 'beta', 'sigma'),
 gg_theme = legend_none()) 

# Returns gtable object, not ggplot object
ggsave('figures/7r_traceplot.svg', arrangeGrob(trace), dpi = 72, height = 6, width = 9) 
```
<center>
![](figures/7r_traceplot.svg)
</center>

```{r}
print(fit_npm, pars = c('alpha', 'beta', 'sigma'), 
                         probs=c(0.025, 0.50, 0.975), 
                         digits_summary=3)
```




### 7) Posterior predictive check to evaluate model performance
How well did the model perform? We can perform posterior predictive checks to see if sampled distributions from the fitted model can approximate the density of `SalesPrice` in the training data. If the model performs well, it should be able to retrodict the density of the data used to train the model. The blue lines are the predictions drawn from the joint posterior distribution compared with the observed density of the target $y$ variable.

```{r eval=FALSE}
color_scheme_set(scheme = "blue")
yrep <- extract(fit_npm)[["y_hat"]]
samples <- sample(nrow(yrep), 300)
ppc_dens_overlay(as.vector(df$log_sales_price_z), yrep[samples, ]) + 
  ggsave('figures/8r_posterior_predictive_check.png', dpi = 300, height = 6, width = 9)  
```

<center>
![](figures/8r_posterior_predictive_check.svg)
</center>


Reversing the data transformations gives back the posterior predictive checks on the natural scale (rescale $y$ and exponentiate `log(SalesPrice)` to get back `SalesPrice`:
```{r}
# Take 100 samples of posterior predictive checks and revert back to natural scale
ppc <- yrep[samples, ] %>% 
  t() %>% 
  apply(., MARGIN  = 2, FUN = function(x) exp((x * sd(df$log_sales_price)) + mean(df$log_sales_price))) %>% 
  as.data.frame() %>%
  pivot_longer(everything())

# Plot densities
ggplot(ppc, aes(value)) + 
  geom_density(aes(group = name), colour = "lightblue") + 
  geom_density(data = (df %>% select(SalePrice) %>% rename(value = SalePrice)), colour = 'black') +
  theme(legend.position="none", axis.text.y=element_blank()) +
  labs(title = 'Posterior predictive checks - Black: observed SalePrice\nLight Blue: Posterior Samples') +
  ggsave('figures/9r_posterior_predictive_check_outcomescale.png', dpi = 300, height = 6, width = 9) 
```

<center>
![](figures/9r_posterior_predictive_check_outcomescale.svg)
</center>

Not bad for a simple model. There is definitely room for iteration and improvement.

### Conclusion

The last things we should do is compare the fits of multiple models and evaluate their performance using cross validation. Since we only fitted a single model, I have reserved this for the next post which applies the complete workflow on more complex multilevel/hierarchical models. Model performance can also be evaluated on out of sample test data as well since this is a predictive task (this kaggle competition computes the log RMSE of the out of sample dataset).

This is not an exhaustive review of all the diagnostics and visualiastions that can be performed in a workflow. There are many ways of evaluating model fit and diagnostics that could validate or invalidate the model. I hope this was a useful example of building a simple bayesian model. Please comment below if you have any thoughts or feedback!

```{r}
sessionInfo()-
```
